Positivity-Preserving Adaptive Runge-Kutta Methods

### 2020-03-26 latex/overview.tex
Order of adapted weights, Different versions of Plot
overview.tex:
Added a short text regarding the order of the adapted steps.

Plots:
Changed plots to display the objective function and not the used weights.
More revision of Section 3.

Now the weights are adapted. The results are also plotted in Figure\,\ref{fig:exampleI}, with solid lines. 
The positivity constraint is now fulfilled. A qualitatively correct solution is obtained.
{\bf Need to mention here what the order is when the weights are adapted.}

@@

\begin{enumerate}
    \item Select in advance a set of desirable weight vectors $b^1, b^2, \dots, b^m$
        corresponding to known good methods,
        and restrict the choice of $\bt$ to convex combinations of this set.
    \item Require that $\|bt - b\|_1$ be small, rejecting the step if it is not.
\end{enumerate}

We discuss the first strategy here; the second is deferred to a later section.
Ideally every element of the set of potential weight vectors would correspond to
a method of the same order as the original method.   Due to linearity of the
order conditions, any linear combination of such weights would also yield a 
method of the same order.  On the other hand, 
it is natural to include a weight vector corresponding to the forward Euler method
(for explicit methods) or backward Euler method (for implicit methods), since
these two methods guarantee positivity (unconditionally for backward Euler
and conditionally for forward Euler).
We can formulate an LP using the first strategy (restriction to a convex set)
as follows.  Let $B$ denote the matrix with columns $b^1, b^2, \dots, b^m$ and let
$g\in{\mathbb R}^m$.  The LP is then

\LPblock{
(Convex adaptation) Given $F, p$, and $b$, find $g$ that minimizes $||\tilde{b} - b||_1$ subject to
\begin{subequations}
\begin{align}
\bt & = Bg \\
0 & \le g_i \le 1 \\
u^{n+1}&=u^n+\dt F \tilde{b} \geq 0 \label{eq:direct_pos}\\
Q_\pt\tilde{b} &=r_\pt \label{eq:direct_Order}.
\end{align}
\end{subequations}
}

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_direct}
        \caption{Free adaptation}
        \label{fig:b_direct}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_convex}
        \caption{Convex adaptation}
        \label{fig:b_convex}
    \end{subfigure}
    \caption{Graphical representation of the two different approaches to adapt the weights}\label{fig:b_space}
\end{figure}

@@

@@

\subsection{Convex Adaptation of Existing Methods}\label{sec:convex}
For some test problems (e.g. the Heat Equation solved with an implicit method of higher order), large changes occur, even though a norm of $\Delta b$ is used as an objective function. A numerical example is given in Section\,\ref{sec:Numeric_Results}.
It is known that the backward Euler method leads to a positive solution for any $\dt$ \cite{hundsdorfer_numerical_2003}. The backward Euler is still only first Order but shows more exact results than a method of higher Order that is distorted by the adaption of the weights, at least at some steps. 
A possible method to fix this is to generate the new b based on a convex adaptation of embedded methods. 
The new weights are defined as 
\begin{equation}\label{eq:b_convex_def}
\tilde{b} = \sum_i^d g_i b_i = Bg
\end{equation}
where $B=\left[b_1,\cdots,b_d \right]$ is a matrix containing the embedded methods.  
The LP is defined as
\LPblock{
\begin{subequations}
\begin{align}
 u^{n+1}&=u^n+\dt K B g \geq 0 \label{eq:Convex_pos}  \\
 0 \leq g_i &\leq 1  \forall {i \in \{1, \cdots d \}} \quad \sum_{i=1}^d g_i = 1 \label{eq:Convex_cvx} \\
 \min_{g} &\left(\sum_i^d w_i g_i \right) \label{eq:Convex_obj}
\end{align}
\end{subequations}
}

\subsubsection{Constraints}
At first a set of embedded methods of the RKM $b_1,\cdots,b_d$ is needed. 
If all the used weights complie with the order conditions, the new method is also of the same order.
This is straightforward using
$$ Q b = Q (\chi b_{\alpha} + (1-\chi) b_{\beta}) = \chi Q  b_{\alpha} + (1-\chi) Q b_{\beta} = \chi r + (1-\chi) r = r$$
$b_{\alpha}$ and $b_{\beta}$ are the methods that are combined to $b$ using the parameter $\chi$. The Matrix $Q$ and the vector $r$ are the order conditions as defined in Section\,\ref{sec:OrderCond}.
To ensure that \eqref{eq:b_convex_def} yields to a convex adaptation the constraints \eqref{eq:Convex_cvx} are enforced for the parameters $g_1,\cdots,g_d$.
The positivity constraint changes to \eqref{eq:Convex_pos}

\subsubsection{Objective Function}

As an objective function \eqref{eq:Convex_obj}
is used.  The $w_1,\cdots,w_d$ are weighting factors that determine which is the preferred $b_i$ to use.

Because the optimal solution of the unconstrained problem has to be the original weights the weight corresponding to $b$ should be the smallest $w$. This weight is simply set to $w_{b} = 0$. For the other weights $w_i = \frac{1}{\mathrm{Order} \{b_i\}}$ is used. This prefers embeded methods of higher order.

@@

With the parameters $a=\num{1e-2} $ and $ d=\num{1e-6}$. 
The PDE is simulated on the domain $x = [0,1]$ with $N=100$ points. Periodic boundary conditions are used. 
The advection part is semidiscreticed using a first order upwind scheme and the diffusion part is semidiscreticed using a 3-point-scheme. This leads to a positivity preserving system of ODEs.  
The computation is done using the EX3 method with free adaptation. 
As step size control a PI-Control from\,\cite{hairer_solving_1996} is used. The Error was estimated using \eqref{eq:Err}. The Tolerance was set to $Tol = 0.01$.  
The final time is $t_{end} = 50$. 
The simulation required 264 steps. Of these, 72 required an adaptation of the weights. 
The solutions for $t=9,t=18,t=27$ and $t=50$ are plotted in Figure\,\ref{fig:Sol_ADP}. 
In Figure\,\ref{fig:sol_ADP18} it can be seen that the reaction occurs in a small interfaces. 
Outside of this regions quantities are close to $0$. Therefore, it is very likely that negative values occur in the numerical approximation. 
In Figure\,\ref{fig:sol_ADP27} it can be seen that at $T=27$ the two reaction interfaces merged. Afterwards the reaction stops and the behavior is mainly controlled by the advection and diffusion part.

>>>>

Now the weights are adapted. The adapted weights are of 4th order. The results are also plotted in Figure\,\ref{fig:exampleI}, with solid lines. 
The positivity constraint is now fulfilled. A qualitatively correct solution is obtained.

@@

\begin{enumerate}
    \item Select in advance a set of desirable weight vectors $b^1, b^2, \dots, b^m$
        corresponding to known good methods,
        and restrict the choice of $\bt$ to convex combinations of this set.
    \item Require that $\|\bt - b\|_1$ be small, rejecting the step if it is not.
\end{enumerate}

We discuss the first strategy here; the second is deferred to a later section.
Ideally every element of the set of potential weight vectors would correspond to
a method of the same order as the original method.   Due to linearity of the
order conditions, any linear combination of such weights would also yield a 
method of the same order.  On the other hand, 
it is natural to include a weight vector corresponding to the forward Euler method
(for explicit methods) or backward Euler method (for implicit methods), since
these two methods guarantee positivity (unconditionally for backward Euler
and conditionally for forward Euler).
We can formulate an LP using the first strategy (restriction to a convex set)
as follows.  Let $B$ denote the matrix with columns $b^1, b^2, \dots, b^m$ and let
$g\in{\mathbb R}^m$.  The LP is then as follows:

\LPblock{
(Convex adaptation) Given $F, p$, and $b$, find $g$ that minimizes $||\tilde{b} - b||_1$ subject to
\begin{subequations}
\begin{align}
\bt & = Bg \\
0 & \le g_i \le 1 \\
u^{n+1}&=u^n+\dt F \tilde{b} \geq 0 \label{eq:direct_pos}.
\end{align}
\end{subequations}
}

Note that we do not need to impose the order conditions here, since they will be
satisfied by each of the methods and thus (by linearity) by the modified
method.  The order of the modified method will in general be equal to the lowest
order among the component methods.

@@

\begin{figure}
    \centering
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_direct}
        \caption{Free adaptation}
        \label{fig:b_direct}
    \end{subfigure}
    \begin{subfigure}[b]{0.45\textwidth}
        \input{Image_b_convex}
        \caption{Convex adaptation}
        \label{fig:b_convex}
    \end{subfigure}
    \caption{Graphical representation of the two different approaches to adapt the weights, for
    a two-stage method.}\label{fig:b_space}
\end{figure}

@@

{\bf Need to also discuss relaxing the order conditions}

@@

@@

With the parameters $a=\num{1e-2} $ and $ d=\num{1e-6}$. 
The PDE is simulated on the domain $x = [0,1]$ with $N=100$ points. Periodic boundary conditions are used. 
The advection part is semidiscreticed using a first order upwind scheme and the diffusion part is semidiscreticed using a 3-point-scheme. This leads to a positivity preserving system of ODEs.  
The computation is done using the EX3 method with free adaptation. 
As step size control a PI-Control from\,\cite{hairer_solving_1996} is used. The Error was estimated using \eqref{eq:Err}. The Tolerance was set to $Tol = 0.01$.  
The final time is $t_{end} = 50$. 
The simulation required 264 steps. Of these, 72 required an adaptation of the weights. 
All adapted weights are still of 3rd order.
The solutions for $t=9,t=18,t=27$ and $t=50$ are plotted in Figure\,\ref{fig:Sol_ADP}. 
In Figure\,\ref{fig:sol_ADP18} it can be seen that the reaction occurs in a small interfaces. 
Outside of this regions quantities are close to $0$. Therefore, it is very likely that negative values occur in the numerical approximation. 
In Figure\,\ref{fig:sol_ADP27} it can be seen that at $T=27$ the two reaction interfaces merged. Afterwards the reaction stops and the behavior is mainly controlled by the advection and diffusion part.
